Towards typed repositories of proofs

### 2010-06-18 papers/MIPS2010/abstract.tex
On allège encore: abstract court sans technique
git-svn-id: http://gasp.dev.regis-gianas.com/svn/trunk@113 403e655e-980a-47bd-8c60-1ee5348bb6b7

%\renewcommand{\remplan}[1]{}
%\renewcommand{\remtext}[1]{}

@@

\section{Motivations}

@@

Practical efforts in the formalization of mathematical results have
naturally led to the questions of how to manage large repositories of
proofs, i.e. what is the \emph{daily workflow} of users of a proof
assistant. How does one elaborate a formal proof? What kind of \emph{a
  posteriori} modification is he prone to doing? What do these
modifications imply on the validity of the whole edifice or, the other
way around, how do one rely on existing work to build up new results?
Many of these questions remain largely unanswered, but the tendency
seems to be to adapt existing methods coming from software
development, as illustrated for example by the introduction of
modules, file-based scripts and separate compilation in proof
assistants like \textsf{Coq}~{\cite{CoqDocWeb}} or
\textsf{Matita}~{\cite{Asperti06userinteraction}}, the use of
dependency management tools (\textsf{Make}) or version control system
(\textsf{GIT}, \textsf{Subversion}) to build and manage versions of a
project. Both for the development of proofs or programs, these tools
attempt to cope with the fact that most of a mathematician or
programmer's time is actually spent \emph{editing}, not
\emph{writing}.

@@

We will describe here some of the possible directions to develop a
tool to analyze the impact of change through types. It involves at its
core a typed description language for repositories, described in
\S\ref{kernel}. In the first iteration of this project, we focus on a
static, or data-driven model for repositories; in \S\ref{rw}, we
quickly review related approaches, and hint a possible generalization
in \S\ref{fw}.

\section{A core language to describe typed repositories}
\label{kernel}

@@

\subsection{How to describe a development?}

@@

\subsection{Formalization}

We describe the state of a repository at a given moment by a type in
the following syntax. Well-typed types in the repository meta-language
guarantee that it contains only well-typed proofs in the object
language.

\paragraph{Syntax} There are three syntactic categories: applicative
terms $a$, types $t$ and environments $\Gamma$.

\begin{align*}
 \gdecl{a}{x \gor a\ x } \\
 \gdecl{t}{a \gor \sort \gor (x:t)\cdot t \gor (x=a:t)\cdot t} \\
 \gdecl{\Gamma}{\cdot \gor \Gamma[x:t] \gor \Gamma[x=a:t]}
\end{align*}

Following the standard syntax, we denote by $(x:t)\cdot u$ the
dependent product (elsewhere written with a $\Pi$). The new construct
$(x=a:t)\cdot u$ is a definition binder, whose value is $a$. We write
as usual $(x_1:t_1)(x_2:t_2)\ldots(x_n:t_n)\cdot t$ for
$(x_1:t_1)\cdot(x_2:t_2)\cdot\ldots\cdot(x_n:t_n)\cdot t$, $t \to u$
for $(x:t)\cdot u$ where $x\notin u$, and $(x\ y : t)\cdot u$ for
$(x:t)(y:t)\cdot u$. The lookup in the environment is written either
$\Gamma(x):t$, denoting ``$\Gamma$ contains a declaration $[x:t]$
\emph{or} a definition $[x=a:t]$'', or $\Gamma(x)=a:t$ denoting
``$\Gamma$ contains a definition $[x=a:t]$''. Capture-avoiding
renaming of variables in a term $t$ is written $t\subst{x}{y}$ for
``replace all occurences of $x$ by $y$'' and is defined the usual way.

\paragraph{Typing} Typing is parameterized by:
\begin{itemize}
  \item$\mathcal S$ a set of sorts \sort,
  \item$\mathcal A \in \mathcal S^2$ the axioms of the system,
  \item$\mathcal R \in \mathcal S^3$ the allowed products.
\end{itemize}
It relies on two distinct judgments:
\begin{itemize}
\item $\Gamma\vdash a : t$ ($a$ has type $t$ in $\Gamma$),
\item $\Gamma\vdash t : \sort$ (the repository $t$ is well-typed of
  sort $s$ in $\Gamma$).
\end{itemize}
The rules are:
\begin{mathpar}
  \infer[Axiom]{
  }{\Gamma\vdash\sort_1:\sort_2}
  \quad(\sort_1,\sort_2)\in\mathcal A
  \and
  \infer[Init]{
    \Gamma(x):t
  }{\Gamma\vdash x:t}
  \and
  \infer[Prod]{\Gamma\vdash t:\sort_1 \and 
    \Gamma[x:t]\vdash u:\sort_2
  }{\Gamma\vdash(x:t)\cdot u : \sort_3}
  \quad(\sort_1,\sort_2,\sort_3)\in\mathcal R
  \and
  \infer[Let1]{\Gamma\vdash a:t \and
    \Gamma\vdash t:\sort_1 \and
    \Gamma[x=a:t]\vdash u:\sort_2
  }{\Gamma\vdash(x=a:t)\cdot u : \sort_3} 
  \quad (x=a:t)\notin\Gamma
  \and
  \infer[Let2]{\Gamma\vdash t:\sort_1 \and
    \Gamma\vdash u:\sort_2
  }{\Gamma\vdash(x=a:t)\cdot u : \sort_3} 
  \quad(x=a:t)\in\Gamma
  \and
  \infer[App]{
    \Gamma\vdash a:(y:t)\cdot u \and \Gamma(x):t
    }{\Gamma\vdash a\ x : u\subst{x}{y}}
  \and
  \infer[Skip1]{
    \Gamma\vdash a:(y=b:t)\cdot u \and \Gamma(x)=b:t
    }{\Gamma\vdash a : u\subst{x}{y}} \and
  \infer[Skip2]{
    \Gamma[x=b:t]\vdash a:(y=b:t)\cdot u
    }{\Gamma\vdash a : u\subst{x}{y}}
\end{mathpar}

The first four rules are standard. As application is only performed on
variables, the \textsc{App} rule looks up directly in the
environment. Rules \textsc{Let1} and \textsc{Skip1} show the
verification and the use of the equality binder in case the definition
is not already known. Rule \textsc{Let2} and \textsc{Skip2} allow
incremental typing in our system: if the definition was already known
in the environment, we do not need to recheck it. If $\cdot\vdash
t:\sort$, then the repository $t$ is well-typed.

\paragraph{Example}

Here is a well-typed repository describing the simply-typed
$\lambda$-calculus and constructing two $\lambda$-terms sharing some
sub-derivations: 
{\small
\begin{align}&
(\nat : *).
(o : \nat).
(s : \nat \to \nat).
(type : *).
(base : type).
(arr : type \to type \to type). \\&
(env : *).
(nil : env).
(cons : type \to env \to env). \\&
(term : *).
(var : \nat \to term).
(lam : term \to term).
(app : term \to term \to term). \\&
(lookup : env \to \nat \to type \to *).
(lookup_o : (e : env).(a : type).(f = cons\ a\ e : env). lookup\ e\ o\ a). \\&
(lookup_s : (e : env).(a b : type).(n : \nat).(sn = s\ n : \nat).(f =
cons\ a\ e : env). lookup\ e\ n\ b \to lookup\ f\ sn\ b). \\&
(of : env \to term \to type \to *). \\&
(of_v : (e : env).(a : type).(n : \nat).(v = var\ n : term).
          lookup\ e\ n\ a \to of\ e\ v\ a). \\&
(of_l : (e : env).(t : term).(a\ b: type).(c = arr\ a\ b : type).
          (u = lam\ t : term).(f = cons\ a\ e : env).
          of\ f\ t\ b \to of\ e\ u\ c). \\&
(of_a : (e : env).(t\ u: term).(a\ b: type).
          (c = arr\ a\ b : type).(v = app\ t\ u : term).
          of\ e\ t\ c \to of\ e\ u\ a \to of\ e\ v\ b). \\&
(1 = s\ o : \nat).(v_0 = var\ o : term).(v_1 = var\ 1 : term).
(t_3 = lam\ v_0 : term).(ch_0 = lam\ t_3 : term).\\&
(t_1 = app\ v_0\ v_1 : term).(t_2 = lam\ t_1 : term).
(t_4 = lam\ t_2 : term).
(bb = arr\ base\ base : type). \\&
(e_0 = cons\ bb\ nil : env).
(e_1 = cons\ base\ e0 : env).
(d_0 = lookup_o\ e_1\ base : lookup\ e_1\ o\ base). \\& %        (* a→a, a ⊢ 0 : a *)
(d_1 = of_v\ e_1\ base\ o\ d0 : of\ e_1\ v_0\ base). %	   (* a→a, a ⊢ 0 : a *)
(d_2 = of_l\ e_0\ v_0\ base\ base d_1 : of\ e_0\ t_3\ bb). \\& %   (* a→a ⊢ λ0 : a → a *)
(bbb = arr\ bb\ bb : type).
(d_3 = of_l\ nil\ t_3\ bb\ bb\ d_2 : of\ nil\ ch_0\ bbb). \\& %  (* ⊢ λλ0 : (a→a)→(a→a) *)
(g_1 = lookup_o\ e_0\ bb : lookup\ e_0\ o\ bb).
(g_2 = lookup_s\ e_0\ base\ bb\ o\ g_1 : lookup\ e_1\ 1\ bb). \\&
(g_3 = of_v\ e_1\ bb\ 1\ g_2 : of\ e_1\ v_1\ bb).	          % (* a→a, a ⊢ 1 : a→a *)
(t_5 = app\ v_1\ v0 : term). \\&
(g_4 = of_a\ e_1\ v_1\ v_0\ base\ base\ g_3\ d_1 : of\ e1\ t5\ base). % (* ... ⊢ 1 0 : a *)
(t_6 = lam\ t_5 : term). \\&
(g_5 = of_l\ e_0\ t_5\ base\ base\ g_4 : of\ e_0\ t_6\ bb).  %  (* a→a ⊢ λ(1 0) : a→a *)
(t_7 = lam\ t_6 : term). \\&
(g_6 = of_l\ nil\ t_6\ bb\ bb\ g_5 : of\ nil\ t_7\ bbb). *	 %  (* ⊢ λλ(1 0) : (a→a)→a→a  *)
\end{align}
}

\section{Related work}
\label{rw}

@@

\section{Further work}
\label{fw}

\remtext{Extension vers une version constructiviste?}

\remtext{Parler du bootstraping?}

\section{Conclusion}

>>>>

\renewcommand{\remplan}[1]{}
\renewcommand{\remtext}[1]{}

@@

\paragraph{Motivations}

@@

Practical efforts in the formalization of mathematical results have
naturally led to the questions of how to manage large repositories of
proofs, i.e. what is the \emph{daily workflow} of users of a proof
assistant. How does one elaborate a formal proof? What kind of \emph{a
  posteriori} modification is he prone to doing? What do these
modifications imply on the validity of the whole edifice or, the other
way around, how do one rely on existing work to build up new results?
Many of these questions remain largely unanswered, but the tendency
seems to be to adapt existing methods coming from software
development, as illustrated for example by the introduction of
modules, file-based scripts and separate compilation in proof
assistants like \textsf{Coq} ({\cite{CoqDocWeb}}) or
\textsf{Matita}~{\cite{Asperti06userinteraction}}, the use of
dependency management tools (\textsf{Make}) or version control system
(\textsf{GIT}, \textsf{Subversion}) to build and manage versions of a
project. Both for the development of proofs or programs, these tools
attempt to cope with the fact that most of a mathematician or
programmer's time is actually spent \emph{editing}, not
\emph{writing}.

@@

We will describe some of the possible directions to develop a tool to
analyze the impact of change through types. It involves at its core a
typed description language for repositories, and is strongly related
to incremental type-checking: only differences between versions are
type-checked and not the entire development. In the first iteration of
this project, we focus on a static, or data-driven model for
repositories inspired by the repository model of \textsf{GIT}.

\paragraph{A core language to describe typed repositories}

@@

@@

We describe the state of a repository at a given moment by a type in
the following syntax. Well-typed types in the repository meta-language
guarantee that it contains only well-typed proofs in the object
language.
\begin{align*}
 \gdecl{a}{x \gor a\ x } \\
 \gdecl{t}{a \gor \sort \gor (x:t)\cdot t \gor (x=a:t)\cdot t} \\
 \gdecl{\Gamma}{\cdot \gor \Gamma[x:t] \gor \Gamma[x=a:t]}
\end{align*}

% \label{kernel}

% The kernel of our system is a type-checker algorithm for a typed
% meta-language. In this language, we will declare both the syntax of
% the object (proof-)language and its typing rules, and define pieces of
% syntax (our proofs) and their derivations. Describing transformations
% among syntax objects is done by sharing common subterms or
% subderivations.

% \subsection{How to describe a development?}

% Representing syntax and logics is nicely done in a \emph{logical
%   framework} like LF (\cite{harper1993framework}): both the syntactic
% elements and the typing derivations can sit in the same tree
% structure, and both can be rechecked at the same time, thanks to
% dependent types. For the purpose of incremental type checking though,
% our needs are a little bit different: first, we need to record, that
% is to name all intermediate values of our developments, so as to be
% able to address and reuse them multiple times. Secondly, we need to
% make sure that those intermediate values (sub-terms) are not recorded
% twice, so as to not type-check them twice: we are looking to represent
% syntactic and typing objects as a DAG rather than a tree.

% Our system will have these properties w.r.t. LF:
% \begin{itemize}
% \item In this first iteration of the project, we do not need
%   computations to take place within our DAGs. Our syntax will then be
%   restricted to product types and applications;
% \item Every term should be a flat application of variables, so that we
%   don't introduce compound terms without naming them;
% \item Finally we need a way to record intermediate definitions: we
%   introduce a new kind of binder, the equality binder. We maintain the
%   invariant that not two same definition can sit in the context while
%   typing.
% \end{itemize}

% \subsection{Formalization}

% We describe the state of a repository at a given moment by a type in
% the following syntax. Well-typed types in the repository meta-language
% guarantee that it contains only well-typed proofs in the object
% language.

% \paragraph{Syntax} There are three syntactic categories: applicative
% terms $a$, types $t$ and environments $\Gamma$.

% \begin{align*}
%  \gdecl{a}{x \gor a\ x } \\
%  \gdecl{t}{a \gor \sort \gor (x:t)\cdot t \gor (x=a:t)\cdot t} \\
%  \gdecl{\Gamma}{\cdot \gor \Gamma[x:t] \gor \Gamma[x=a:t]}
% \end{align*}

% Following the standard syntax, we denote by $(x:t)\cdot u$ the
% dependent product (elsewhere written with a $\Pi$). The new construct
% $(x=a:t)\cdot u$ is a definition binder, whose value is $a$. We write
% as usual $(x_1:t_1)(x_2:t_2)\ldots(x_n:t_n)\cdot t$ for
% $(x_1:t_1)\cdot(x_2:t_2)\cdot\ldots\cdot(x_n:t_n)\cdot t$, $t \to u$
% for $(x:t)\cdot u$ where $x\notin u$, and $(x\ y : t)\cdot u$ for
% $(x:t)(y:t)\cdot u$. The lookup in the environment is written either
% $\Gamma(x):t$, denoting ``$\Gamma$ contains a declaration $[x:t]$
% \emph{or} a definition $[x=a:t]$'', or $\Gamma(x)=a:t$ denoting
% ``$\Gamma$ contains a definition $[x=a:t]$''. Capture-avoiding
% renaming of variables in a term $t$ is written $t\subst{x}{y}$ for
% ``replace all occurences of $x$ by $y$'' and is defined the usual way.

% \paragraph{Typing} Typing is parameterized by:
% \begin{itemize}
%   \item$\mathcal S$ a set of sorts \sort,
%   \item$\mathcal A \in \mathcal S^2$ the axioms of the system,
%   \item$\mathcal R \in \mathcal S^3$ the allowed products.
% \end{itemize}
% It relies on two distinct judgments:
% \begin{itemize}
% \item $\Gamma\vdash a : t$ ($a$ has type $t$ in $\Gamma$),
% \item $\Gamma\vdash t : \sort$ (the repository $t$ is well-typed of
%   sort $s$ in $\Gamma$).
% \end{itemize}
% The rules are:
% \begin{mathpar}
%   \infer[Axiom]{
%   }{\Gamma\vdash\sort_1:\sort_2}
%   \quad(\sort_1,\sort_2)\in\mathcal A
%   \and
%   \infer[Init]{
%     \Gamma(x):t
%   }{\Gamma\vdash x:t}
%   \and
%   \infer[Prod]{\Gamma\vdash t:\sort_1 \and 
%     \Gamma[x:t]\vdash u:\sort_2
%   }{\Gamma\vdash(x:t)\cdot u : \sort_3}
%   \quad(\sort_1,\sort_2,\sort_3)\in\mathcal R
%   \and
%   \infer[Let1]{\Gamma\vdash a:t \and
%     \Gamma\vdash t:\sort_1 \and
%     \Gamma[x=a:t]\vdash u:\sort_2
%   }{\Gamma\vdash(x=a:t)\cdot u : \sort_3} 
%   \quad (x=a:t)\notin\Gamma
%   \and
%   \infer[Let2]{\Gamma\vdash t:\sort_1 \and
%     \Gamma\vdash u:\sort_2
%   }{\Gamma\vdash(x=a:t)\cdot u : \sort_3} 
%   \quad(x=a:t)\in\Gamma
%   \and
%   \infer[App]{
%     \Gamma\vdash a:(y:t)\cdot u \and \Gamma(x):t
%     }{\Gamma\vdash a\ x : u\subst{x}{y}}
%   \and
%   \infer[Skip1]{
%     \Gamma\vdash a:(y=b:t)\cdot u \and \Gamma(x)=b:t
%     }{\Gamma\vdash a : u\subst{x}{y}} \and
%   \infer[Skip2]{
%     \Gamma[x=b:t]\vdash a:(y=b:t)\cdot u
%     }{\Gamma\vdash a : u\subst{x}{y}}
% \end{mathpar}

% The first four rules are standard. As application is only performed on
% variables, the \textsc{App} rule looks up directly in the
% environment. Rules \textsc{Let1} and \textsc{Skip1} show the
% verification and the use of the equality binder in case the definition
% is not already known. Rule \textsc{Let2} and \textsc{Skip2} allow
% incremental typing in our system: if the definition was already known
% in the environment, we do not need to recheck it. If $\cdot\vdash
% t:\sort$, then the repository $t$ is well-typed.

% \paragraph{Example}

% Here is a well-typed repository describing the simply-typed
% $\lambda$-calculus and constructing two $\lambda$-terms sharing some
% sub-derivations: 
% {\small
% \begin{align}&
% (\nat : *).
% (o : \nat).
% (s : \nat \to \nat).
% (type : *).
% (base : type).
% (arr : type \to type \to type). \\&
% (env : *).
% (nil : env).
% (cons : type \to env \to env). \\&
% (term : *).
% (var : \nat \to term).
% (lam : term \to term).
% (app : term \to term \to term). \\&
% (lookup : env \to \nat \to type \to *).
% (lookup_o : (e : env).(a : type).(f = cons\ a\ e : env). lookup\ e\ o\ a). \\&
% (lookup_s : (e : env).(a b : type).(n : \nat).(sn = s\ n : \nat).(f =
% cons\ a\ e : env). lookup\ e\ n\ b \to lookup\ f\ sn\ b). \\&
% (of : env \to term \to type \to *). \\&
% (of_v : (e : env).(a : type).(n : \nat).(v = var\ n : term).
%           lookup\ e\ n\ a \to of\ e\ v\ a). \\&
% (of_l : (e : env).(t : term).(a\ b: type).(c = arr\ a\ b : type).
%           (u = lam\ t : term).(f = cons\ a\ e : env).
%           of\ f\ t\ b \to of\ e\ u\ c). \\&
% (of_a : (e : env).(t\ u: term).(a\ b: type).
%           (c = arr\ a\ b : type).(v = app\ t\ u : term).
%           of\ e\ t\ c \to of\ e\ u\ a \to of\ e\ v\ b). \\&
% (1 = s\ o : \nat).(v_0 = var\ o : term).(v_1 = var\ 1 : term).
% (t_3 = lam\ v_0 : term).(ch_0 = lam\ t_3 : term).\\&
% (t_1 = app\ v_0\ v_1 : term).(t_2 = lam\ t_1 : term).
% (t_4 = lam\ t_2 : term).
% (bb = arr\ base\ base : type). \\&
% (e_0 = cons\ bb\ nil : env).
% (e_1 = cons\ base\ e0 : env).
% (d_0 = lookup_o\ e_1\ base : lookup\ e_1\ o\ base). \\& %        (* a→a, a ⊢ 0 : a *)
% (d_1 = of_v\ e_1\ base\ o\ d0 : of\ e_1\ v_0\ base). %	   (* a→a, a ⊢ 0 : a *)
% (d_2 = of_l\ e_0\ v_0\ base\ base d_1 : of\ e_0\ t_3\ bb). \\& %   (* a→a ⊢ λ0 : a → a *)
% (bbb = arr\ bb\ bb : type).
% (d_3 = of_l\ nil\ t_3\ bb\ bb\ d_2 : of\ nil\ ch_0\ bbb). \\& %  (* ⊢ λλ0 : (a→a)→(a→a) *)
% (g_1 = lookup_o\ e_0\ bb : lookup\ e_0\ o\ bb).
% (g_2 = lookup_s\ e_0\ base\ bb\ o\ g_1 : lookup\ e_1\ 1\ bb). \\&
% (g_3 = of_v\ e_1\ bb\ 1\ g_2 : of\ e_1\ v_1\ bb).	          % (* a→a, a ⊢ 1 : a→a *)
% (t_5 = app\ v_1\ v0 : term). \\&
% (g_4 = of_a\ e_1\ v_1\ v_0\ base\ base\ g_3\ d_1 : of\ e1\ t5\ base). % (* ... ⊢ 1 0 : a *)
% (t_6 = lam\ t_5 : term). \\&
% (g_5 = of_l\ e_0\ t_5\ base\ base\ g_4 : of\ e_0\ t_6\ bb).  %  (* a→a ⊢ λ(1 0) : a→a *)
% (t_7 = lam\ t_6 : term). \\&
% (g_6 = of_l\ nil\ t_6\ bb\ bb\ g_5 : of\ nil\ t_7\ bbb). *	 %  (* ⊢ λλ(1 0) : (a→a)→a→a  *)
% \end{align}
% }

\paragraph{Related work}
\label{rw}

@@

% \section{Further work}
% \label{fw}

% \remtext{Extension vers une version constructiviste?}

% \remtext{Parler du bootstraping?}

% \section{Conclusion}
